---
title: "Human_graph"
output: html_document
date: "2026-09-17"
---

# Load libraries

```{r, warning=FALSE, error=FALSE}

setwd("~/Human_graph")
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(readxl)
library(writexl)
library(igraph)
library(ggraph)
library(visNetwork)
```

# Load and roughly filter the dataset

```{r}
# Select SNP filtering column based on dataset 

```

```{r, warning=FALSE}

v66_p1_compatibility_HO_aadr_PUB_anno <- readxl::read_xlsx("aadr_v66_HOCompatibility_unfiltered.xlsx")

# REMOVE DUPLICATES BY INDIVIDUAL ID COLUMN AND RETAIN THE ONE WITH MORE SNPs
df_filt_nodupl <- v66_p1_compatibility_HO_aadr_PUB_anno %>% 
  group_by(`Individual ID`) %>%
  filter(`SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)` == max(`SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)`, na.rm = TRUE)) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(`Persistent Genetic ID`) 

# FILTER FOR > 5000 SNPs
# REMOVE CRITICAL INDIVIDUALS
df_filtsnps_nodupl_noCRIT <- df_filt_nodupl %>% 
  filter(
    `SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)`>= 5000,
    !`Persistent Genetic ID` %in% c(27759,27760,28478),
    !`ASSESSMENT` %in% c("CRITICAL", "MERGE_CRITICAL", "PROVISIONAL_CRITICAL")
    )

# STORE REMOVED INDIVIDUALS FOR LATER CHECK
removed_ind <- df_filt_nodupl[
    df_filt_nodupl$`SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)`< 5000 |
    df_filt_nodupl$`Persistent Genetic ID` %in% c(27759,27760,28478) |
    df_filt_nodupl$`ASSESSMENT` %in% c("CRITICAL", "MERGE_CRITICAL", "PROVISIONAL_CRITICAL"),
]
  
# REMOVE UNFILTERED TABLES
rm(df_filt_nodupl)
rm(v66_p1_compatibility_HO_aadr_PUB_anno)
```

Separate relatives from related\
Individuals with no relatives **should** have NA, ".." or "" in the family relations column...

```{r}

# CHECK HOW MANY UNRELATED
sum(is.na(df_filtsnps_nodupl_noCRIT$`Family relations`)) 
table(df_filtsnps_nodupl_noCRIT[(nchar(df_filtsnps_nodupl_noCRIT$`Family relations`) <5), "Family relations"])
# 172 NA and 19827 ".."

unrelated_symbols <- c("", "..")

# STORE "UNRELATED" SAMPLES
unrelat_S <- df_filtsnps_nodupl_noCRIT %>%
  filter(`Family relations` %in% unrelated_symbols | is.na(`Family relations`))

# STORE "RELATED"
related_noS <- df_filtsnps_nodupl_noCRIT %>%
  filter(!(`Family relations` %in% unrelated_symbols | is.na(`Family relations`)))
  
```

# Create relationship table (edges)

```{r}

edges <- related_noS %>%
  select(`Individual ID`, `Family relations`) %>%
  
  # Split comma-separated relationships into separate rows
  separate_rows(`Family relations`, sep = ",") %>%
  
  # Remove whitespace
  mutate(`Family relations` = str_trim(`Family relations`)) %>%

  # Extract the rel_deg
  mutate(
    rel_deg = as.numeric(
      str_extract(`Family relations`, "\\d+(?:\\.\\d+)?(?=d)")
    )
  ) %>%
  
  # Extract the two individual IDs
  mutate(
    id1 = str_match(
      `Family relations`,
      "d.*?:([^:]+):([^,]+)"
    )[, 2],
    
    id2 = str_match(
      `Family relations`,
      "d.*?:([^:]+):([^,]+)"
    )[, 3]
  ) %>%
  
  select(id1, id2, rel_deg)
```

## Quality check

```{r}
edges <- edges %>%
  mutate(
    id1 = str_trim(id1),
    id2 = str_trim(id2)
  ) %>%
  filter(
    !is.na(id1),
    !is.na(id2),
    id1 != "",
    id2 != "",
    !is.na(rel_deg)
  )
```

## Add weights to relationships

```{r}

# rel_deg REFERS TO RELATIONSHIP
edges$rel_deg <- as.numeric(edges$rel_deg)

# ADDING WEIGHTS FOR RELATIONSHIP (customize it)
edges <- edges %>%
  mutate(weight = case_when(
    rel_deg == 0.5 ~ 2,
    rel_deg == 1 ~ 3,
    rel_deg == 1.5 ~ 2.5,
    rel_deg == 2 ~ 2,
    rel_deg == 2.5 ~ 1.5,
    rel_deg == 3 ~ 1,
    TRUE ~ NA_real_
  ))

edges$weight <- as.numeric(edges$weight)

```

## Remove duplicated nodes (undirected graph)

```{r}

edges <- edges %>%
  mutate(
    from = pmin(id1, id2),
    to   = pmax(id1, id2)
  ) %>%
  distinct(from, to, .keep_all = TRUE) %>%
  select(-from, -to)

```

## Fix problems

```{r}

# STORE NODES NAMES
nodes_names <- unique(c(edges$id1, edges$id2))

# CREATE NODES
nodes <- data.frame(
  id = related_noS$`Individual ID`,
  SNP = related_noS$`SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)`,
  stringsAsFactors = FALSE
)

length(unique(c(edges$id1,edges$id2))) # 3482
length(unique(nodes$id)) # 3333

# PROBLEM!
# we have more nodes from the edges tables than expected...
# some individuals from the "unrelated" group are mentioned in edges so they are related
# let's fix this and put them in the right place

unique(c(edges$id1[edges$id1 %in% unrelat_S$`Individual ID`],
         edges$id2[edges$id2 %in% unrelat_S$`Individual ID`]))
# 20 individuals from the unrelated

edges_1 <- edges[((edges$id1 %in% nodes$id) & (edges$id2 %in% nodes$id)),]
length(unique(c(edges_1$id1,edges_1$id2))) # 3262
# REMOVING 220 NODES

e1 <- unique(c(edges$id1, edges$id2)[!(c(edges$id1,edges$id2) %in% c(edges_1$id1,edges_1$id2))])
sum(e1 %in% nodes$id)                     # 71 are present in the nodes
sum(e1 %in% unrelat_S$`Individual ID`)    # 20 should have been unrelated but are not
sum(e1 %in% removed_ind$`Individual ID`)  # 129 were removed from the initial filter

# save them here, move them and redo filters and quality controls
e2 <- e1[e1 %in% unrelat_S$`Individual ID`]

# REMOVE FROM UNRELATED THE RELATED ONES
unrelat_S <- unrelat_S[!unrelat_S$`Individual ID` %in% e2,] # removed from unrelated

# ADD RELATED IN THRE RELATED DF
related_noS <- rbind(related_noS, df_filtsnps_nodupl_noCRIT[df_filtsnps_nodupl_noCRIT$`Individual ID` %in% e2,])  

# RECREATE NODES DF
nodes <- data.frame(
  id = related_noS$`Individual ID`,
  SNP = related_noS$`SNPs hit on autosomal targets (Computed using easystats on Compatibility_HO snpset)`,
  stringsAsFactors = FALSE)

length(unique(c(edges$id1,edges$id2)))
length(unique(c(edges_1$id1,edges_1$id2)))

```

## Reconstruct the edges

```{r}

edges_2 <- related_noS %>%
  select(`Individual ID`, `Family relations`) %>%
  
  # Split comma-separated relationships into separate rows
  separate_rows(`Family relations`, sep = ",") %>%
  
  # Remove whitespace
  mutate(`Family relations` = str_trim(`Family relations`)) %>%

  # Extract the rel_deg
  mutate(
    rel_deg = as.numeric(
      str_extract(`Family relations`, "\\d+(?:\\.\\d+)?(?=d)")
    )
  ) %>%
  
  # Extract the two individual IDs
  mutate(
    id1 = str_match(
      `Family relations`,
      "d.*?:([^:]+):([^,]+)"
    )[, 2],
    
    id2 = str_match(
      `Family relations`,
      "d.*?:([^:]+):([^,]+)"
    )[, 3]
  ) %>%
  
  select(id1, id2, rel_deg)


# QUALITY CHECK
# removing NA rows resulted from related individuals which have ".." or NA in family rel
edges_2 <- edges_2 %>%
  mutate(
    id1 = str_trim(id1),
    id2 = str_trim(id2)
  ) %>%
  filter(
    !is.na(id1),
    !is.na(id2),
    id1 != "",
    id2 != "",
    !is.na(rel_deg)
  )

## add weights
edges_2$rel_deg <- as.numeric(edges_2$rel_deg)
edges_2 <- edges_2 %>%
  mutate(weight = case_when(
    rel_deg == 0.5 ~ 2,
    rel_deg == 1 ~ 3,
    rel_deg == 1.5 ~ 2.5,
    rel_deg == 2 ~ 2,
    rel_deg == 2.5 ~ 1.5,
    rel_deg == 3 ~ 1,
    TRUE ~ NA_real_
  ))

edges_2$weight <- as.numeric(edges_2$weight)


# remove duplicated nodes (undirected graph)
edges_2 <- edges_2 %>%
  mutate(
    from = pmin(id1, id2),
    to   = pmax(id1, id2)
  ) %>%
  distinct(from, to, .keep_all = TRUE) %>%
  select(-from, -to)

# end of ploblem fix
```

# Check and filter some edges

```{r}

length(unique(c(edges_2$id1, edges_2$id2))) # 3482 vertices in edges
nodes_names1 <- unique(nodes$id)            # 3353 vertices in nodes 

# e4 = nodes in edges_2 that are not stored in nodes
e3 <- unique(c(edges_2$id1, edges_2$id2))
e4 <- setdiff(e3, nodes_names1)

sum(e4 %in% removed_ind$`Individual ID`)
# all the 129 nodes not present in nodes are already removed from the initial filtering
```

We have 3482 nodes in edges, 129 are in "removed_ind" (x), 3353 are fine (o)

| ID1 | ID2 |                                                 |
|-----|-----|-------------------------------------------------|
| x   | x   | case (1): remove the edge, there should be none |
| o   | o   | case (2): fine, keep                            |
| o   | x   | case (3): remove edge                           |
| x   | o   | case (4): remove edge                           |

Filtering to keep only case 2, from case 3 and 4 we will still have the nodes stored

```{r}
edges_3 <- edges_2[
  (edges_2$id1 %in% nodes_names1) &
  (edges_2$id2 %in% nodes_names1),]            

length(unique(c(edges_3$id1, edges_3$id2))) 
# 3290 vertices with connections, 3353 total nodes
```

At this point we have:

-   unrelated samples unrelated_S

-   nodes that have family relationships

-   nodes that have NO relationships (because the other sample has been filtered out)

# Set selection

Greedy approximation to select a large independent set.

until there are edges in the graph; calculate the degree of each node and remove the one with highest degree. If multiple nodes share the maximum degree, delete the one with less SNPs. It deletes the vertex from the graph. When every remaining node has degree 0, it returns the remaining nodes.

```{r}

# GRAPH
g <- graph_from_data_frame(
  edges_3,
  directed = FALSE,
  vertices = nodes
)
```

```{r}

# ALGORITHM
greedy_unrelated_set <- function(g) {
  g_work <- g  # work on a copy so we don't touch the original graph
  
  while (ecount(g_work) > 0) {
    degs <- degree(g_work)
    max_deg <- max(degs)
    tied <- names(degs)[degs == max_deg]
    
    if (length(tied) > 1) {
      # among tied nodes, pick the one with the lowest SNP value
      snp_vals <- V(g_work)[tied]$SNP
      worst <- tied[which.min(snp_vals)]
    } else {
      worst <- tied
    }
    
    g_work <- delete_vertices(g_work, worst)
  }
  
  V(g_work)$name  # remaining nodes = largest unrelated set (greedy)
}

# test
set.seed(123)
unrelated_nodes <- greedy_unrelated_set(g)
length(unrelated_nodes)
head(unrelated_nodes)
```

```{r}
# best_set <- c()
# for (i in 1:100) {
#   set.seed(i)
#   candidate <- greedy_unrelated_set(g)
#   if (length(candidate) > length(best_set)) best_set <- candidate
# }

best_set <- unrelated_nodes
length(best_set)
is_ivs(g, best_set)
```

# Store independent IDs

```{r}
ind_IDs <- c(best_set, unrelat_S$`Individual ID`)
length(ind_IDs)


final_ind <- df_filtsnps_nodupl_noCRIT[df_filtsnps_nodupl_noCRIT$`Individual ID` %in% ind_IDs,]  

```

# Filter based on political entity

```{r}
pol_ent_list <- unique(final_ind$`Political Entity`)
pol_ent_list

# Choose the desired political entities
selected_pol_ent <- c("Italy", "Greece") 
selecte_ind <- final_ind[final_ind$`Political Entity` %in% selected_pol_ent,]

```
